Safe Transferable Regions

### 2016-10-14 paper/abstract.tex
proofs and space compression

% Recent work in high-performance systems written in managed languages
% (such as Java or C\#) has shown that garbage-collection can be a
% significant performance bottleneck.
% A class of these systems, focused on big-data, create numerous large
% data structures with well-defined lifetimes.
%
There is increasing interest in alternative memory management techniques
that strike a balance between the convenience of garbage collection and
the better performance that manual memory management enables, without
compromising safety.
%
In this paper, we present a language and a memory
management scheme based on user-managed memory regions (called
\emph{transferable regions}) that allow programmers to exploit the
knowledge of data structures' lifetimes to improve performance.

Manual memory management is susceptible to the usual perils of
dangling pointers. A key contribution of this paper is a
refinement-based region type system that ensures the memory safety of
C\# programs in the presence of transferable regions.
We complement our type system with a type inference algorithm.
This eliminates the need for programmers to write region annotations on types.
We establish the soundness of the type inference algorithm and
show that the key steps of constraint-generation and constraint-solving
in the type inference algorithm are complete.

>>>>

% Recent work in high-performance systems written in managed languages
% (such as Java or C\#) has shown that garbage-collection can be a
% significant performance bottleneck.
% A class of these systems, focused on big-data, create numerous large
% data structures with well-defined lifetimes.
%
There is increasing interest in alternative memory management techniques
that strike a balance between the convenience of garbage collection and
the better performance that manual memory management enables, without
compromising safety.
%
In this paper, we present a language and a memory management scheme based
on user-managed memory regions (called \emph{transferable regions})
that allow programmers to exploit knowledge of data structures' lifetimes
to improve performance.

Manual memory management is susceptible to the usual perils of
dangling pointers. A key contribution of this paper is a
refinement-based region type system, for a C\#-like typed object-oriented
language with higher-order functions and generics, that ensures the
memory safety of programs in the presence of transferable regions.
We complement our type system with a type inference algorithm,
which eliminates the need for programmers to write region annotations on types.
The cornerstone of our inference algorithm is a novel constraint solver that
performs abduction in a partial-order constraint domain to infer weakest solutions
to recursive constraints.
We establish soundness and partial completeness results for the type inference algorithm:
in particular, the key steps of constraint-generation and constraint-solving
in the type inference algorithm are complete.

### 2016-10-14 paper/typeinference-proofs.tex
proofs and space compression

\begin{theorem}
\label{thm:constraint-solver-completeness}
Let $C = \consOf{q}$.
If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
% If $\satC$ is satisfiable, then the constraint solver will return some assignment.
\end{theorem}

\begin{proof}
If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
$\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
Consider any such $\rho$.
Let $\soln$ be a satisfying assignment for $\satC$.
Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
$\pi \in S_\rho$ and $S_\rho$ cannot be empty.
If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
The result follows.
\end{proof}

@@

% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
% 
% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
% It is easy to show that we take care of these.
% 
% The more difficult case are validity constraints, involving either region variables
% ($\rho$) or predicate variables ($\varphi$) or both.
% 
% We use this property to show that it is okay to ignore validity constraints
% in $\satC$ containing region variables. For every
% region variable $\rho$ unified with a constant $\pi$, and for every validity
% constraint containing $\rho$, we will have a corresponding validity constraint
% containing $\pi$ instead.
% 
% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
% to ignore these constraints.
% 
% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
% us convert these into simpler constraints that do not have $\varphi$ on the
% antecedent or into a constraint that is directly incorporated into definition
% of $\predSubstFn$.
\end{proof}

>>>>

\begin{theorem}
\label{thm:constraint-solver-completeness}
Let $C = \consOf{q}$.
If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
\end{theorem}

\begin{proof}
We prove the theorem by contradiction.
Assume that $\satC$ is satisfiable.
We will show that the precondition for the first case in the definition of $\textsc{Solve}$
holds and, hence, $\solveCon{C}$ cannot be $\textsc{None}$.

Since $\satC$ is satisfiable, $\groundC$ is trivially valid.
We just need to show that $\thesolnR$ satisfies $\rhoC$.
Recall that $\rhoC$ is the subset of $C$ of well-formedness constraints
on region variables.
So, we just need to show for any constraint $\rho \in \rhoenv$ in $C$ that
$\thesolnR(\rho) \in \rhoenv$.
Since we have assumed that $\satC$ is satisfiable, there exists some $\soln$ that satisfies $\satC$
and, in particular, the constraint $\rho \in \rhoenv$.
Let $\pi$ denote $\soln(\rho)$. Thus, we have $\pi \in \rhoenv$.

Let $\hat{\pi}$ denote $\thesolnR(\rho)$. If $\pi$ and $\hat{\pi}$ are the same region
constant, then $\hat{\pi}\ \in \rhoenv$ and the proof is complete.
%
Otherwise, it follows from the definition of $\thesoln$, that we have some constraint
$\isvalid{\varphi \conj \phi}{\rho = \hat{\pi}}$ in $\satC$.
Since $\soln$ satisfies this constraint, we have
$\isvalid{\solnP(\varphi) \conj \phi}{\pi = \hat{\pi}}$.
We can show that $\{ \pi, \hat{\pi} \} \subseteq \predDeltaMap(\varphi)$
(as in the proof of Lemma~\ref{lemma:two-bindings}).
We can show, by induction over the constraint-generation rules, that if $\Delta$
includes some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
Hence, it follows that $\hat{\pi} \in \Delta$.
% for some predicate variable $\varphi$.

% If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
% $\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
% Consider any such $\rho$.
% Let $\soln$ be a satisfying assignment for $\satC$.
% Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
% From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
% Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
% Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
% If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
% $\pi \in S_\rho$ and $S_\rho$ cannot be empty.
% If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
% in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
% We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
% some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
% It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
% The result follows.
\end{proof}

@@
